---
title: "Health & Economic Outcomes of Severe Weather Events"
author: "C. McBride"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Synopsis

The U.S. National Oceanic and Atmospheric Administration's (NOAA) Severe Weather dataset tracks extreme weather events in the United States and includes in the summative data for each event estimates of casualties and property damage. This report will focus on these two features of the dataset by compiling similar events types for the purpose of ranking the resulting categories by consequential human casualties and material damage. 

The data themselves required extensive cleaning and reformatting. The wide-ranging weather event types and naming variations were gathered into general categories based on similar characteristics such as temperature, percipitation type and storm variety. The results I think supported more meaningful and interpretable descriptions of the data.

While all event categories registered significant impacts, the events that caused the greatest loss in both terms of life and property were hurricanes, tornados and flooding. Data recording conventions further complicated the data. Some event types are characterized by multiple weather features, such as hurricanes the effects of which are often recorded in the dataset by weather features such as wind or flood. To compensate for this and accurately reflect the casualties and damages caused by a single storm events other variables were harnessed to group the data according to event. (9 sentences) 


## Data Processing & Exploratory Analysis

The data, in particular the focus variable of interest `EVTYPE` which describes the event type, were inconsistent in their formatting. To correct these inconsistencies in preparation for plotting and addressing the research questions, a series of evaluative and data cleaning steps were implemented.

__Step 1:__ The requisite libraries are loaded to facilitate analysis and data cleaning. 

```{r Get Packages, message=FALSE, warning=FALSE}
# load packages
library(dplyr)
library(ggplot2)
library(gtable)
library(knitr)
library(lubridate)
library(readr)
library(reshape2)
library(scales)
library(stringi)
library(xtable)
```

__Step 2:__ The dataset is loaded into the working environment and all possible values of `EVTYPE` are catalogued.

```{r Load Data, message=FALSE, warning=FALSE}
# load compressed data as tibble
noaa <- read_csv("noaa_extreme_weather.csv.bz2")

# get domain of severe weather events
unq <- unique(noaa$EVTYPE)
length(unq)
head(unq)
```

__Step 3:__ After informally cataloging the large number of weather event types (using `table(noaa$EVTYPE)`), it was clearly necessary not only to clean up the many errors and disparities in the data but also gather the event type data into categories so that similar events could be examined and assessed as categories. The reasoning here was that similar events often develop under similar conditions and share feature and damage profiles, for example flooding from heavy rain and flooding from storm surge. To accomplish this, several grouping functions were implemented that rely on regular expressions.

```{r Clean Data, message=FALSE, warning=FALSE}
## clean `EVTYPE` data
# format all levels to uppercase (reduces unique levels by almost 100)
noaa[, "EVTYPE"] <- toupper(noaa$EVTYPE)

# remove summary categories, create dataframe copy
noaa_clean <- noaa[!(stri_detect_regex(noaa$EVTYPE, "SUMMARY")),]

## simplify categories
# find and rename all extreme winter weather events (except snow) under one name
cold <- stri_detect_regex(as.character(noaa_clean$EVTYPE), 
                          "(BLIZZARD|WINT|FREEZE|COLD|ICE|SLEET|ICY)")
noaa_clean[cold, "EVCAT"] <- "COLD"; rm(cold)

# gather all variations of names for hail events under one level name
hail <- stri_detect_regex(as.character(noaa_clean$EVTYPE), "HAIL")
noaa_clean[hail, "EVCAT"] <- "HAIL"; rm(hail)

# all flood related extreme weather events under one level name
flood <- stri_detect_regex(as.character(noaa_clean$EVTYPE),
                           "(FLOOD|TSUNAMI|FLD|HIGH TIDE|SURF|SEICHE|SURGE)")
noaa_clean[flood, "EVCAT"] <- "FLOOD"; rm(flood)

# all major oceanic storms under one level name
hurricane <- stri_detect_regex(as.character(noaa_clean$EVTYPE),
                               "(HURRICANE|TYPHOON|TROPICAL|SURGE)")
noaa_clean[hurricane, "EVCAT"] <- "HURRICANE"; rm(hurricane)

# all thunderstorm and extreme rain under one level name
tstorm <- stri_detect_regex(as.character(noaa_clean$EVTYPE), 
                            "(THUN| TSTM|RAIN|PRECIP)")
noaa_clean[tstorm, "EVCAT"] <- "THUNDERSTORM"; rm(tstorm)

# all extreme heat and drought events under one name
heat_dry <- stri_detect_regex(as.character(noaa_clean$EVTYPE), 
                              "(DROUGHT|DRY|HOT|WARM|HEAT)")
noaa_clean[heat_dry, "EVCAT"] <- "HEAT_DRY"; rm(heat_dry)

# all tornado type events under one level name
tornado <- stri_detect_regex(as.character(noaa_clean$EVTYPE), 
             ("TORNAD|SPOUT|WHIRL|FUNNEL|ROTATING WALL CLOUD|DUST DEVIL"))
noaa_clean[tornado, "EVCAT"] <- "TORNADO"; rm(tornado)

# all extreme wind events (except tornadoes) under one level name
wind <- stri_detect_regex(as.character(noaa_clean$EVTYPE), 
                          "(WIND|BURST|WHIRL)")
noaa_clean[wind, "EVCAT"] <- "WIND"; rm(wind)

# all variations of snow under one level name
snow <- stri_detect_regex(as.character(noaa_clean$EVTYPE), "SNOW")
noaa_clean[snow, "EVCAT"] <- "SNOW"; rm(snow)

# all variations of wildfire descriptions under one level name
fire <- stri_detect_regex(as.character(noaa_clean$EVTYPE), "(FIRE|SMOKE)")
noaa_clean[fire, "EVCAT"] <- "FIRE"; rm(fire)

# all variations of volcano under one level name
volcano <- stri_detect_regex(as.character(noaa_clean$EVTYPE), "VOLCA")
noaa_clean[volcano, "EVCAT"] <- "VOLCANO"; rm(volcano)

# all variations of volcano under one level name
fog<- stri_detect_regex(as.character(noaa_clean$EVTYPE), "FOG")
noaa_clean[fog, "EVCAT"] <- "FOG"; rm(fog)
```

__Step 4:__ Event types that were infrequent or didn't fit within the defined broader categories were eliminated, and event types and categories that were infrequently represented (n < 500) over the entire timespan of the data were not included. 

```{r}
# assign EVTYPE value to EVCAT for those event types that weren't grouped under
# a larger category in previous step
nas <- is.na(noaa_clean$EVCAT)
noaa_clean$EVCAT[nas] <- noaa_clean$EVTYPE[nas]

# remove most infrequent categories
freq_evts <- table(noaa_clean$EVCAT)[table(noaa_clean$EVCAT) > 500]
noaa_clean <- noaa_clean[noaa_clean$EVCAT %in% names(freq_evts),]

# changes to feature classes
noaa_clean[, "EVCAT"] <- as.factor(noaa_clean$EVCAT)
```

__Step 5:__ The numeric codes for `STATE__` and `COUNTY` are formatted and combined into a [FIPS county code][3] for possible geographic plotting by county.

```{r Generate FIPS codes}
# COUNTY code formatting function, adds requisite zeros
noaa_clean <- noaa_clean%>%
  mutate(FIPS=as.integer(STATE__ * 1000 + COUNTY))

```

__Step 6:__ To facilitate subsetting the data on the particular dates to capture exceptional events such as large hurricanes, the date fields `BGN_DATE` and `END_DATE` are transformed form character objects to Date objects.

```{r Format Dates, warning=FALSE, message=FALSE}
# transform date variable from character class to Date class
noaa_clean <- noaa_clean%>%
  mutate(BGN_DATE=as.Date(BGN_DATE, format="%m/%d/%Y %H:%M:%S"),
         END_DATE=as.Date(END_DATE, format="%m/%d/%Y %H:%M:%S"))
```

__Step 7:__ To facilitate geographic plotting, a couple of transformations are made to the data. To be able to plot with choropleths, a variable with the full state name in all lower case is added to the data.

```{r Add full state name, message=FALSE, warning=FALSE}
# Use STATE to create feature with full state name
noaa_clean <- noaa_clean%>%
  mutate(STATENAME=tolower(state.name[match(STATE, state.abb)]))
  
```

In addition, the coordinate values, `LONGITUDE` and `LATITUDE` are cleaned and formatted. Although the data includes events from US territories and the two states outside the contiguous US, the data was further subsetted on the lower 48 states. 


```{r}
## Prepare geographic plot
# correct coordinate format
# format longitude values
frmt_lons <- function(x){ 
  x2 <- x%/%100 + x%%100/100
  if(x2<0){
    return(x2)
  }else{
    return(x2*-1)
  }
}

# format latitude values
frmt_lats <- function(x){ 
  x%/%100 + x%%100/100
}

# Apply functions for coordinal data
noaa_clean$LONGITUDE <- sapply(noaa_clean$LONGITUDE, frmt_lons) 
noaa_clean$LATITUDE <- sapply(noaa_clean$LATITUDE, frmt_lats)

# Further subset coordinate values against bounding box of continental US
noaa_clean <- subset(noaa_clean, STATE %in% state.abb)
```


__Step 8:__ As a preliminary formatting step for examining material damage, an integer estimate of the damages caused by each event is calculated using the coefficient and exponent variable. Observations with anomalous values (e.g. `PROGDMGEXP`="?")  are removed from the dataset.


```{r}
## Calculate damages with PROPDMG and CROPDMG with their exponent variables
# deal with NA's in exponent columns
noaa_clean[is.na(noaa_clean$PROPDMGEXP), 'PROPDMGEXP'] <- "0"
noaa_clean[is.na(noaa_clean$CROPDMGEXP), 'CROPDMGEXP'] <- "0"

# Set up look-up list for transforming exponent variable to numerical
dmg_exps <- list("K"=3, "M"=6, "B"=9, "m"=6, "h"=2, "H"=2, "k"=3)

# Deal with anomalous characters in PROPDMGEXP and CROPDMGEXP
noaa_clean[noaa_clean$PROPDMGEXP %in% c("+", "?", "-"),
           'PROPDMGEXP'] <- "0"
noaa_clean[noaa_clean$CROPDMGEXP %in% c("+", "?", "-"),
           'CROPDMGEXP'] <- "0"

# Transform letter exponent to number
noaa_clean$PROPDMGEXP <- sapply(noaa_clean$PROPDMGEXP,
                                function(e){
                                  ifelse(e %in% names(dmg_exps), 
                                         dmg_exps[[e]], as.numeric(e))
                                })

noaa_clean$CROPDMGEXP <- sapply(noaa_clean$CROPDMGEXP,
                                function(e){
                                  ifelse(e %in% names(dmg_exps), 
                                         dmg_exps[[e]], as.numeric(e))
                                })
# Perform exponent calculations, assign full cost value back to PROPDMG, CROPDMG
noaa_clean <- noaa_clean%>%
  mutate(PROPDMG_TOT=PROPDMG*10^PROPDMGEXP,
         CROPDMG_TOT=CROPDMG*10^CROPDMGEXP,
         YEAR=year(as.Date(BGN_DATE, format="%m/%d/%Y")))
```

__Step 9:__ Because some the data under `PROPDMG` look extreme, a quick summary to test for outliers.

```{r message=FALSE, warning=FALSE}
# Get quantile summaries of damages by event category
qnt_smry <- noaa_clean%>%
  group_by(EVCAT)%>%
  summarize(min=min(PROPDMG_TOT),
            first=quantile(PROPDMG_TOT, .25),
            median=median(PROPDMG_TOT),
            mean=mean(PROPDMG_TOT),
            third_qrt=quantile(PROPDMG_TOT, .75), 
            max=max(PROPDMG_TOT))

```

The max values will be examined event category by event category to look for outliers or anomalies. In the case of `COLD`, the max value for `PROPDMG_TOT` of $5 billion doesn't seem plausible. The event in question was described as the ["Storm of the Century" or the "Blizzard of 1993"][1] in popular culture and categorized as a category 5 storm meteorologically under which the entire eastern seaboard of the United States experienced extreme weather conditions including record snowfalls in states as far south as Alabama. 

Likewise, the costliest fire turns out not to be a data error. The [Cerro Grande Fire][2] in northern New Mexico along with other fires across the state caused widespread damage worth over a billion dollars.

The damages caused by a severe hail storm in October of 2010 in Phoenix, Arizona caused close to $2 billion in damages.

The maximum value for `THUNDERSTORM` is probably better moved to the flood category. 

The billion dollar damages recorded for the top tornado belonged to a tornado that tore through Joplin, Missouri in May of 2011.

The maximimum value for `WIND` is actually another observation attributable to the ["Storm of the Century"][1]. Many of the other extreme property damages value for this category are attributalbe to hurricanes.

## Results & Conclusions

This dataset, in context of the research questions, look deceptively simple. Many of the weather event types and the generated event categories are closely correlated. Hurricanes, for instance, were often catalogued by their most destructive features high winds and floods caused by rains and storm surges. Analysis by event type or category alone would have overlooked this feature of the data and likely misinterpreted the data, assigning casualty or damage figures to weather characteristics instead of broader event types. To demonstrate this, four especially destructive storms were isolated in the data using date and geographic ranges, and then the observations related to each storm were tabulated by frequency of event category. The results are summarized in the table below:

```{r warning=FALSE, message=FALSE}
## weather event categories associated with single events
# Hurricane Katrina
katrina <- noaa_clean%>%
  filter(BGN_DATE >= as.Date("2005-08-22") &
         END_DATE <= as.Date("2005-09-01") &
         STATE %in% c('AL', 'LA', 'MS', 'KY', 'TN', 'FL') |
         grepl("Katrina", noaa_clean$REMARKS))%>%
  group_by(EVCAT)%>%
  summarise(freq=n())
  
katrina <- cbind(rep("Hurricane Katrina", length(katrina)), 
                     katrina)
names(katrina) <- c("Storm", "Category", "Frequency")

# Joplin Tornado
joplin <- noaa_clean%>%
  filter(BGN_DATE >= as.Date("2011-05-21") &
        END_DATE <= as.Date("2011-05-23") &
        FIPS %in% c(29097, 29145) |
        grepl("Joplin", noaa_clean$REMARKS))%>%
  group_by(EVCAT)%>%
  summarise(freq=n())

joplin <- cbind(rep("Joplin Tornado", nrow(joplin)), 
                     joplin)
names(joplin) <- c("Storm", "Category", "Frequency")

# Storm of Century - Blizzard of 1993
blizzard_93 <- noaa_clean%>%
  filter(BGN_DATE >= as.Date("1993-03-11") &
         BGN_DATE <= as.Date("1993-03-16") &
         STATE %in% c('FL', 'MS', 'AL', 'GA', 'SC', 'NC', 'TN',
                      'KY', 'VA', 'WV', 'IN', 'OH', 'NY', 'PA',
                      'NJ', 'MA', 'ME', 'RI', 'DE', 'CT', 'MD',
                      'VT'))%>%
  group_by(EVCAT)%>%
  summarise(freq=n())  

blizzard_93 <- cbind(rep("Blizzard of 1993", nrow(blizzard_93)), 
                     blizzard_93)
names(blizzard_93) <- c("Storm", "Category", "Frequency")



# Hurricane Camille
camille <- noaa_clean%>%
  filter(BGN_DATE >= as.Date("1969-08-15") &
         BGN_DATE <= as.Date("1969-08-21") &
         STATE %in% c('FL', 'OK', 'WV', 'VA', 'DE',
                      'GA', 'LA', 'MS', 'AL', 'TN', 
                      'KY') |
         grepl("Camille", noaa_clean$REMARKS))%>%
  group_by(EVCAT)%>%
  summarise(freq=n())

camille <- cbind(rep("Hurricane Camille", length(camille)), 
                     camille)
names(camille) <- c("Storm", "Category", "Frequency")

perf_storms <- rbind(katrina, joplin, blizzard_93, camille)%>%
               dcast(Category ~ Storm, value.var = "Frequency", fun=sum)

```


```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
# summary table of event categories frequency for four super storms
kable(perf_storms, caption="Observations Attributable to Super-Storms by Event Category")
```

As is clear in the table, the effects of a storm are summarized by storm feature not exclusively by storm category or type. For example, a relatively low proportion of observations attributable to Hurricane Katrina were actually catalogued under hurricane. The table shows that observations were usually listed under the most destructive feature such as wind and flooding. 

Further complicating any direct answer to the research questions, are the distributions of material damage by event category. This data is summarized in the following table:

```{r}
# summaries table of distributions of damages by event category
kable(qnt_smry, caption = "Quantile Distributions of Property Damages by Event Category")
```

The distributions are extremely skewed toward higher values with the third quartile value often being zero. Most weather events catalogued in the dataset don't even result in material damage, and those that do inflict considerable damage. The vast majority of casualties and damages captured by the dataset are caused by a very small proportion of storms. Therefore in response to both research questions, what event types are most harmful to public health and what events have the greatest economic consequences, the answer would be the collective set of features of perfect or super storms of large magnitude and geographic spread as well as historic presidence. 

Plotting the casualties and damages by event category over time visually demonstrates that the vast bulk of human and material damages are owed to super storms.  


```{r}
# group data by EVCAT and get total fatalities, injuries for each category
smry_fatals <- noaa_clean%>%
  filter(YEAR>1990)%>%
  group_by(EVCAT, YEAR)%>%
  summarize(tot_fatals=sum(FATALITIES),
            tot_injuries=sum(INJURIES))%>%
  melt(id.vars=c("EVCAT", "YEAR"), measure.vars=c("tot_fatals", "tot_injuries"))

ggplot(smry_fatals)+
  geom_line(aes(x=YEAR, y=value, color=variable))+
  facet_wrap(~EVCAT)+
  labs(title="Fatalities and Injuries by Event Category")
  theme(axis.text.x = element_text(angle=75, hjust=1))

```

The plots clearly show that historically rare super events in the hurricane, tornado, and flood (often hurricane and tornado correlated) are by far the most dangerous and deadly types of extreme weather. These are seen as spikes in line plots that correspond to these weather events. 





```{r}
# Group data by event category and year, summarize by property and crop totals
smry_dmgs <- noaa_clean%>%
  filter(YEAR > 1990)%>%
  group_by(EVCAT, YEAR)%>%
  summarize(tot_prop=mean(PROPDMG_TOT),
            tot_crop=mean(CROPDMG_TOT))

ggplot(smry_dmgs)+
  geom_line(aes(x=YEAR, y=tot_prop+tot_crop))+
  facet_wrap(~EVCAT)+
  labs(title="Combined Property and Crop Damages by Event Category")+
  theme(axis.text.x = element_text(angle=75, hjust=1))
```


To gain further some insight into how events are categorized in the dataset by their features instead of their broader categorical names, individual storms will be focused on. 

```{r}
# Create vectory of states affected by Hurricane Katrina
af_sts_kat <- c('alabama', 'louisiana', 'mississippi', 
                'kentucky', 'tennessee', 'florida',
                'arkansas', 'illinois', 'missouri', 'virginia',
                'indiana', 'ohio', 'georgia', 'south carolina',
                'north carolina')

# Summarize Hurricane Katrina Damages
katrina_dmgs <- noaa_clean%>%
  filter(BGN_DATE >= as.Date("2005-08-22") &
         BGN_DATE <= as.Date("2005-09-01") &
         STATE %in% c('AL', 'LA', 'MS', 'KY', 'TN', 'FL',
                      'AR', 'IL', 'IN', 'OH', 'GA', 'MO',
                      'SC', 'NC', 'VA') &
         grepl("Katrina", noaa_clean$REMARKS))%>%
  group_by(STATENAME, EVCAT)%>%
  summarise(value=log(sum(PROPDMG_TOT) + sum(CROPDMG_TOT) + 1))

# Rename variables
names(katrina_dmgs)[1] <- "region" 

# Create base map
states_map <- map_data("state", region = af_sts_kat)

# Faceted plot of hurricane damages by event category
ggplot(katrina_dmgs, aes(map_id = region)) + 
    geom_map(data=states_map, map = states_map, color="black", fill=NA)+
    geom_map(aes(fill = value), color="black", map = states_map) +
    scale_fill_continuous(low="yellow", high="red", guide = FALSE) + 
    expand_limits(x = states_map$long, y = states_map$lat)+
    facet_wrap(~EVCAT)+
    labs(Title="Hurricane Katrina State Level Damages by Event Category")+
    theme(axis.ticks=element_blank(), 
          axis.text = element_blank(),
          axis.title = element_blank(),
          axis.line = element_blank(),
          legend.background = element_blank(),
          legend.key = element_blank(),
          panel.grid = element_blank(),
          strip.background = element_blank())


```

The plots show that the damages from Hurricane Katrina in 2005 covered a wide geographic region encompassing several states. The damages were attributed to specific storms weather features in the data, however. This is true for other super storms as well, for example the Blizzard of 1993, which brought a variety of destructive weather to the eastern US.

```{r warning=FALSE, message=FALSE}
# vector of affected states for Blizzard of 1993
af_sts <- c("alabama",  "connecticut", "delaware", "florida", "georgia",
                    "kentucky", "louisiana", "maine", "maryland", "massachusetts",
                    "mississippi", "new hampshire",  "new jersey", "new york",
                    "north carolina", "pennsylvania", "south carolina",
                    "tennessee", "vermont", "virginia", "west virginia")

```

```{r}
# Summarize Blizzard of 1993 Damages
blizzard_93_facet <- noaa_clean%>%
  filter(BGN_DATE >= as.Date("1993-03-11") & 
           BGN_DATE <= as.Date("1993-03-15") &
           !(EVCAT %in% c("HURRICANE", "LIGHTNING")) &
           STATENAME %in% af_sts)%>%
  group_by(STATENAME, EVCAT)%>%
  summarise(value=log(sum(PROPDMG_TOT) + sum(CROPDMG_TOT) + 1))

# Rename variables
names(blizzard_93_facet)[1] <- "region" 

# Create base map
states_map <- map_data("state", region = af_sts)

# Faceted plot of state level damages by event category 
ggplot(blizzard_93_facet, aes(map_id = region)) + 
    geom_map(data=states_map, map = states_map, color="black", fill=NA)+
    geom_map(aes(fill = value), color="black", map = states_map) +
  labs(Title="Blizzard of 1993 State Level Damages by Event Category")+
    scale_fill_continuous(low="yellow", high="red", guide = FALSE) + 
    expand_limits(x = states_map$long, y = states_map$lat)+
    facet_wrap(~EVCAT)+
    theme(axis.ticks=element_blank(), 
          axis.text = element_blank(),
          axis.title = element_blank(),
          axis.line = element_blank(),
          legend.background = element_blank(),
          legend.key = element_blank(),
          panel.grid = element_blank(),
          strip.background = element_blank())


```






## Citations & Links

NOAA Extreme Weather Events Data
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

Data Documentation
https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf


[1]: https://en.wikipedia.org/wiki/1993_Storm_of_the_Century

[2]: https://en.wikipedia.org/wiki/Cerro_Grande_Fire

[3]: https://en.wikipedia.org/wiki/FIPS_county_code